A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis

### 2018-05-04 doc/genesetdp.tex

\begin{abstract}
A prevailing technique to infer function from findings in high-throughput experiments is gene set enrichemnet analysis, where the sets of genes are compared to predefined sets of related genes, so called pathways. As some of such pathways are known to be incomplete, algoritmic efforts have been made to complement them with information from functional association networks, using so called Network Enrichment Analysis (NEA). Traditionally significance of inferences from NEA has been asssigned using network perturbations. Here we instead have designed a dynamic programming  algorithm that can calculte the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engin, NEA.

@@

The significance of the inferences from NEA are assessed by null models that investigate the number of expected random links from the query geneset to the pathway. Previous efforts have settled for varius parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.

Since such methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic. Also, the somewhat inaccessible process used to randomize networks produces statistics that are equaly hard to interpret as the null hypothesis becomes complex. FIXME: Should this be in the dicussion?

@@

This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elemets from $N_{a-1}$ still needed to update subsequent elements. That is, we do not have to copy the dynamic programing matrix in each iteration over $a$, instead we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$. For details see Listing \ref{lst:gensetdp}.

@@

We implemented a Python program that reads network and pathway definition files and scored query sets agains a pathway according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).

@@

In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome. As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed. To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration}. As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.

@@

FIXME: Is it ok to state the method is more accurate if we are testing it agains a different null hypothesis?

@@

Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. Previous implementations of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than null models that relate to model parameters.

>>>>

\begin{abstract}
A prevailing technique to infer function from identifications from molecular biological high-throughput experiments is gene set enrichment analysis, where the identifications are compared to predefined sets of related genes, so-called pathways. As at least some pathways are known to be incomplete in their annotation, algorithmic efforts have been made to complement them with information from functional association networks, with so-called Network Enrichment Analysis (NEA). Traditionally, the significance of inferences from NEA has been assigned using parametric models, that are tuned by fitting parameters by sampling random network perturbations. Here we instead have designed a dynamic programming  algorithm that can calculate the score distribution of NEA scores and makes it possible to assign unbiased mid $p$~values to inferences. We demonstrate that our method obtains a superior statistical calibration as compared to the popular NEA inference engine, BinoX.

@@

The significance of the inferences from NEA is assessed by null models that investigate the number of expected random links from the query gene set to the pathway. Previous efforts have settled for various parametric null distributions, where the parameters are estimated by Monte Carlo-simulations that permute the network definitions by methods with varying degrees of topological conservation.

Since such methods of network randomization are not completely random, they introduce bias in the statistics that may or may not be desirable, as the set of random networks is much larger than the ones of random networks that retain a certain topological characteristic. Also, the somewhat inaccessible process used to randomize networks produces statistics that are equally hard to interpret as the null hypothesis becomes complex.

@@

This enables us to calculate $N$ in-place, at least as long as we update the elements in $N_a$ in a reverse order so that we do not alter elements from $N_{a-1}$ still needed to update subsequent elements. That is, we do not have to copy the dynamic programming matrix in each iteration over $a$, instead, we can add a $D_a$ on to of previous iterations $N_{a-1}(s,c)$, by nested updates over $s \in \{ RQ, RQ-1, \ldots 0 \}$ and $c \in \{ Q, Q-1, \ldots 1 \}$. For details see Listing \ref{lst:gensetdp}.

@@

We implemented a Python program that reads network and pathway definition files and scored query sets against a pathway according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).

@@

In order to test the statistical calibration of our method, we calculated $p$~values for 10000 random picks of gene sets from the investigated genome. As our selection is random and following our null hypothesis, we expected the resulting $p$~values to be uniformly distributed. To test this we plotted the $p$~values against their quantile in Figure \ref{fig:calibration} for $10000$ randomly assembeled queries of $25$ genes from the human genome. As a comparison, we also added a calibration curve for the popular NEA method BinoX\cite{ogris2016novel}.

@@

@@

Here we made an explicit definition of the null hypothesis we employed, which centers on the query, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. Previous implementations of NEA in the literature otherwise use Monte Carlo-simulation to determine parameters for various types parametric distributions, by perturbations of the investigated network. In practice, such perturbations are hard to conduct in a manner that they still resemble the original network, and hence are difficult to use to calculate accurate statistics. In general, it is easier to understand null models that relate to user actions, than more complex null models that relate to model parameters.
